Computer-Assisted Analysis of Emendations of G\"odel's Ontological Proof

### 2014-10-20 Papers/InProgress/LogicAndReligionCongress/Abstract.tex
Improve abstract

\usepackage{url}
\usepackage{commands}

@@

\title[Analysis of Modern Variants of G\"odel's Ontological Argument]
 {Computer-supported Analysis of Modern Variants of G\"odel's Ontological Argument}

@@

\thanks{This work was supported by German National Research Foundation (DFG) under
 grants BE 2501/9-1 and BE 2501/11-1 and by ????}

@@

%----------Author 3
\author[Woltzenlogel-Paleo]{B. Woltzenlogel Paleo}
\address{ 
Favoritenstra{\ss}e 9 \\
Room HA0402 \\
1040 Wien \\
Austria
}
\email{bruno.wp@gmail.com}

@@

%\section{The Modal Collapse}
The axioms in G\"odel's ontological argument 
\cite{GoedelNotes,ScottNotes} (cf. Appendix \ref{apx:Goedel}) imply what is called the modal collapse
\cite{Sobel1987,SobelBook2004} (formula $\varphi \rightarrow \Box
\varphi$, abbreviated as MC, holds for any formula $\varphi$ and not
just for $\exists x. God(x)$ as intended). This fact, which has
recently been confirmed with higher-order automated theorem provers
\cite{C40,J30}, has led to strong criticism of the argument, and
it has therefore stimulated attempts to remedy the problem. Some
authors, including Hajek
\cite{Hajek_der_Mathematiker_2002,Hajek_Magari_and_others_1996}, suggest the use
of cautious instead of full comprehension principles in order to
escape the modal collapse. Others, including Anderson
\cite{anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings}
and Bjordal \cite{bjordal99}, propose to slightly modify
G\"odel's axioms and definitions.

We have studied --- with computer-support --- the emendations proposed
by Anderson and Bjordal (thereby we have retained full comprehension).
More precisely, we have formalized their emendations within the proof
assistant Isabelle/HOL \cite{Isabelle}, and we have then automated
(different variations of) their emendations with the higher-order
reasoners Leo-II \cite{C26}, Satallax \cite{brown2012satallax}, Metis
\cite{Hurd03first-orderproof} and Nitpick \cite{Nitpick}; these systems
are available in Isabelle/HOL.  Our formalizations\footnote{The
  formalizations are available in the subdirectories \url{Anderson}
  and \url{Bjordal} at
  \url{https://github.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/}.}
employ the embedding of higher-order modal logic (HOML) in classical
higher-order logic (HOL) as introduced and employed in previous
work \cite{C40,J30,J23}. In this approach full comprehension is
naturally ``built-in'' since the underlying HOL supports
$\lambda$-abstraction.

We summarize the main results of our experiments, and we start with
Anderson's emendation (cf. Appendix \ref{apx:Anderson}; @Bruno: I do
not yet address T2', L1' and L2' below; we could eventually include those as well?),
which we have analysed for different domain conditions. These
variations were motivated by various comments on Anderson's work in
the literature (@Bruno: maybe we should be more explicit about these
comments).
\begin{itemize}
\item Constant domain semantics\footnote{Cf. Isabelle/HOL source file
    \url{Anderson/Anderson_constant_domain.thy}} (the individual
  domains remain unchanged for all worlds): G\"odel's theorems T1, C
  and T3' can be quickly automated (in logics K and KB, respectively),
  G\"odel's axioms A4 and A5 are proven redundant (the former one in
  logic K4B and the latter one already in K). Moreover, a trivial
  countermodel (consisting of two worlds and two individuals) to MC
  generated by Nitpick (for all mentioned logics); this model also
  proves consistency. 
\item Varying domain semantics\footnote{Cf. Isabelle/HOL source file
    \url{Anderson/Anderson_varying_domain.thy}} (the domains for
  indivduals may vary from world to world, all other domains remain
  constant; various authors have pointed out that varying domain
  quantifiers should be used in Anderson's work; here we consistently
  used varying domain for all individual quantifiers; the modeling of
  the varying domain quantifiers employs an explicit ``Existence''
  relation between indiviudals and worlds as guard): In this setting we obtain the
  same results as above.
\item Mixed variant\footnote{Cf. Isabelle/HOL source files
    \url{Anderson/Anderson_mixed_domain_1/2.thy}} (varying domain
  quantifiers are used only in the definitions of essence and NE;
  cf.~Fitting comments to Anderson in
  \cite{fitting02:_types_tableaus_god}): Also in this setting we
  obtain the same results as above. However, if a varying domain
  quantifier is used only in the definition of NE, then the
  situation slightly changes. Now axiom A5 is no longer provable and
  a countermodel is reported by Nitpick. The remaining results are as before.

%   G\"odel's axioms A4 and A5
%   remain redundant (the former one in logic K4B, the latter already in
%   K). The situation changes when only NE is modeled with a varying
%   domain quantifier. Now, Nitpick reports a countermodel.
\end{itemize}

Our analysis of Bjordal emendation of G\"odel's argument (cf. Appendix \ref{apx:Bjordal}) has produced the following results:
\begin{itemize}
\item Constant domain semantics\footnote{Cf. Isabelle/HOL source files
    \url{Bjordal/Bjordal_A/B/C_constant_domain.thy}}: G\"odel's axiom
  A2, A3 can be quickly automatically derived in logic K from
  Bjordal's definition B:D. A4 can be proved in logic T
  (reflexivity). Proving G\"odel's D1 from B:D is possible in logic
  K4. Vice versa, for the proof that B:D follows from D1, A2, A3 and
  A4 the povers succeed already in logic K. Hence, Bjordal's lemma
  B:L1 holds in logic S4. The provers also show that theorem T3
  follows from B:D, B:A1 and B:A2 already in logic KB. Modal collapse
  does not follow in Bjordal's setting as Nitpick demonstrates with a
  countermodel (consisting of two worlds and one individual). @Bruno: we could eventually also
  also say that for MC1 we need a bigger countermodel?
\item Varying domain semantics\footnote{Cf. Isabelle/HOL source files
    \url{Bjordal/Bjordal_A/B/C_varying_domain.thy}}: The results are
  the same as above. 
\item @Bruno: we could also check the mixed case her; but there doesn't seem to be strong reason for doing so.
\end{itemize}

>>>>

\usepackage{url}
\usepackage{commands}
\usepackage{xcolor}
\usepackage{comment}

@@

\title[Analysis of Emendations of G\"odel's Ontological Argument]
 {Computer-Assisted Analysis of 
 Emendations of G\"odel's Ontological Proof}

@@

\thanks{This work was supported by German National Research Foundation (DFG) under
 grants BE 2501/9-1 and BE 2501/11-1.}

@@

%----------Author 3
\author[Woltzenlogel-Paleo]{B. Woltzenlogel Paleo}
\address{ 
Room HA0402, Favoritenstra{\ss}e 9, 1040 Wien, Austria
}
\email{bruno.wp@gmail.com}

@@

%\section{The Modal Collapse}
The axioms in G\"odel's ontological argument 
\cite{GoedelNotes,ScottNotes} (cf. Appendix \ref{apx:Goedel}) entail what is called the \emph{modal collapse}
\cite{Sobel1987,SobelBook2004}: the formula $\varphi \rightarrow \Box
\varphi$, abbreviated as MC, holds for any formula $\varphi$ and not
just for $\exists x. \mathit{God}(x)$ as intended. This fact, which has
recently been confirmed with higher-order automated theorem provers
\cite{C40,J30}, has led to strong criticism of the argument and stimulated attempts to remedy the problem. Hájek
\cite{Hajek_der_Mathematiker_2002,Hajek_Magari_and_others_1996} proposed the use
of cautious instead of full comprehension principles, and Fitting \cite{fitting02:_types_tableaus_god} suggested that greater care is necessary in the semantics of higher-order quantifiers in the presence of modalities. Others, such as
Anderson
\cite{anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings}
and Bjordal \cite{bjordal99}, proposed slight emendations of
G\"odel's axioms and definitions. They require neither comprehension restrictions nor more complex semantics. Therefore, they are technically simpler to analyze with computer support. We have formalized them using the proof
assistant Isabelle/HOL \cite{Isabelle} together with the automated higher-order
reasoners Leo-II \cite{C26}, Satallax \cite{brown2012satallax}, Metis
\cite{Hurd03first-orderproof}, and Nitpick \cite{Nitpick}.  
Our formalizations\footnote{The
  formalizations are available in the subdirectories \url{Anderson}
  and \url{Bjordal} at
  \url{https://github.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/}.}
employ the embedding of higher-order modal logic (HOML) in classical
higher-order logic (HOL) as introduced in previous
work \cite{C40,J30,J23}. We explored the effect of different domain conditions
 on the provability of lemmas, theorems and even axioms. This was motivated by a controversy between Hájek and Anderson regarding the redundancy of some axioms in Anderson's emendation. In \emph{constant domain semantics}, the individual domains are the same in all possible worlds. In \emph{varying domain semantics}, the domains may vary from world to world. This variation is technically encoded with the help of an existence relation expressing which individuals actually exist in each world. Quantifiers are then uniformly formalized as \emph{actualist quantifiers} (i.e. guarded by the existence relation). Our main results are summarized in the following sections.

\section{Anderson's Emendation (cf. Appendix \ref{apx:Anderson})}

\BWP{Bruno: I do
not yet address T2', L1' and L2' below; we could eventually include those as well?}

\BWP{Bruno: maybe we should be more explicit about these
comments}

\paragraph{Constant domain semantics}
%\footnote{Cf. Isabelle/HOL source file \url{Anderson/Anderson_constant_domain.thy}} 
     G\"odel's theorems T1, C
  and T3' can be quickly automated (in logics K and KB, respectively),
  G\"odel's axioms A4 and A5 are proven redundant (the former one in
  logic K4B and the latter one already in K). Moreover, a trivial
  countermodel (consisting of two worlds and two individuals) to MC
  generated by Nitpick (for all mentioned logics); this model also
  shows the consistency of the axioms.

\paragraph{Varying domain semantics}
%\footnote{Cf. Isabelle/HOL source file \url{Anderson/Anderson_varying_domain.thy}} 
: In this setting we obtain the
  same results as above.

\paragraph{Mixed variant}
%\footnote{Cf. Isabelle/HOL source files \url{Anderson/Anderson_mixed_domain_1/2.thy}} 
(varying domain
  quantifiers are used only in the definitions of essence and NE;
  cf.~Fitting's comments to Anderson in
  \cite{fitting02:_types_tableaus_god}): Also in this setting we
  obtain the same results as above. However, if a varying domain
  quantifier is used only in the definition of NE, then the
  situation changes slightly. Now axiom A5 is no longer provable and
  a countermodel is reported by Nitpick. The remaining results are as before.

%   G\"odel's axioms A4 and A5
%   remain redundant (the former one in logic K4B, the latter already in
%   K). The situation changes when only NE is modeled with a varying
%   domain quantifier. Now, Nitpick reports a countermodel.

\section{Bjordal's Emendation (cf. Appendix \ref{apx:Bjordal})}

\paragraph{Constant domain semantics}
%\footnote{Cf. Isabelle/HOL source files \url{Bjordal/Bjordal_A/B/C_constant_domain.thy}}: 
    G\"odel's axiom
  A2, A3 can be quickly automatically derived in logic K from
  Bjordal's definition B:D. A4 can be proved in logic T
  (reflexivity). Proving G\"odel's D1 from B:D is possible in logic
  K4. Conversely, the proof that B:D follows from D1, A2, A3 and
  A4 is possible already in logic K. Hence, Bjordal's lemma
  B:L1 holds in logic S4. The provers also show that theorem T3
  follows from B:D, B:A1 and B:A2 already in logic KB. Modal collapse
  does not follow in Bjordal's setting as Nitpick demonstrates with a
  countermodel (consisting of two worlds and one individual).

  \BWP{@Bruno: we could eventually also
  also say that for MC1 we need a bigger countermodel?}
\paragraph{Varying domain semantics}
%\footnote{Cf. Isabelle/HOL source files \url{Bjordal/Bjordal_A/B/C_varying_domain.thy}}
: The results are
  the same as above.

\section{Conclusions}

Anderson's emendation (cf. Appendix \ref{apx:Anderson}; ,
which we have analysed for different domain conditions. These
variations were motivated by various comments on Anderson's work in
the literature.

In this approach full comprehension is
naturally ``built-in'' since the underlying HOL supports
$\lambda$-abstraction.
